Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-07 manuscript.tex
manuscript edits, sim code

Finally, we tested for population substructure between the two sex chromosome races to control for the possibility that hidden substructure in our sampled populations could affect estimates of diversity. To do this, we conducted a phylogenetic analysis of sex-linked sequences from all populations in our study. The analysis was conducted on an alignment of X- and Y-linked genes from \textit{R. hastatulus}, with orthologous autosomal sequences from the non-dioecious but closely related outgroup species \textit{R. bucephalophorus} used to root the tree (Figure S2). The phylogeny was inferred using the Neighbor-Joining method \citep{saitou1987neighbor}, with evolutionary distances computed using Maximum Composite Likelihood \citep{tamura2011mega5}. The inferred sequence phylogeny revealed strong support for Y-linked genes from the $XY_{1}XY_{2}$ ("North Carolina") race being paraphyletic (98 percent bootstrap support), with samples from two populations (hereafter the SC sub-clade) forming a monophyletic group, and samples from Florida and Georgia (hereafter the FL sub-clade) also forming a monophyletic group, but more closely related to the XY (Texas) race (Figure S2). In light of this evidence for substructure, we used the inferred phylogeny as the basis for estimating X and Y diversity.

@@

To consider the joint effects of biased sex ratios and variance in male reproductive success, our neutrally-expected levels of diversity were calculated assuming no variance in female mating (e.g., $C^2_{f}=0$) and we considered a range of values for $\sigma_{m}$: 0 (no variance), 1/2 (Poisson/2) 1 (Poisson distribution of offspring), 2 (Poisson*2), and 3 (Poisson*3). Assuming the level of neutral polymorphism is given by the product of the mutation rate and the effective population size, $\theta=4N_{e}\mu$ \citep{watterson1975,kimura1984}, Figure 1 shows the predicted X/A and Y/A diversity ratios as a function of the sex ratio and variance in male reproductive success. Note that we have only considered variance in male reproductive success because this is most likely to occur in wind pollinated plant like \textit{R. hastatulus}, where males produce copious amounts of pollen and compete to fertilize uniovulate females.

@@

\subsection*{Simulations of positive and purifying selection}
To investigate whether our observed level of Y-chromosome diversity could be explained by background selection, we conducted forward time simulations, modeling both purifying and positive selection on the Y chromosome using the software SFSCODE \citep{hernandez2008flexible}. To make the simulation output comparable to our data, we initialized simulations with an empirically estimated value of $\theta$ (from autosomes), and sampled 8 Y chromosomes per simulation from 24 populations. To consider the effects of selection against deleterious alleles (purifying selection), we sampled selective coefficients from a gamma distribution with shape parameter $k=0.1$, mean selection coefficient $s=-0.01$, neutral mutation rate $\mu = 0.005$, deleterious mutation rate $U=0.01$ (diploid). For positive selection simulations, we used \X [parameters here]. We ran 1000 replicates for each model and we calculated the average $\theta_{sim}$ over all simulation output to obtain the simulated Y/A ratio of neutral diversity under the both models (purifying and positive selection), taking into account unequal male and female population sizes. We compared simulated diversity output to our empirical data (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{sim}$ was not significantly different from $\theta_{observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supplementary Material.

>>>>

Finally, we tested for population substructure between the two sex chromosome races to control for the possibility that hidden substructure in our sampled populations could affect estimates of diversity. To do this, we conducted a phylogenetic analysis of sex-linked sequences from all populations in our study. The analysis was conducted on an alignment of X- and Y-linked genes from \textit{R. hastatulus}, with orthologous autosomal sequences from the non-dioecious but closely related outgroup species \textit{R. bucephalophorus} used to root the tree (Figure S2). The phylogeny was inferred using the Neighbor-Joining method \citep{saitou1987neighbor}, with evolutionary distances computed using Maximum Composite Likelihood \citep{tamura2011mega5}. The inferred sequence phylogeny revealed strong support for Y-linked genes from the $XY_{1}XY_{2}$ ("North Carolina") race being paraphyletic (98 percent bootstrap support), with samples from two populations (hereafter the SC sub-clade) forming a monophyletic group, and samples from Florida and Georgia (hereafter the FL sub-clade) also forming a monophyletic group, but more closely related to the XY (Texas) race (Figure S2). In light of this evidence for population substructure, we used the inferred phylogeny as the basis for estimating X and Y diversity.

@@

To consider the joint effects of biased sex ratios and variance in male reproductive success, our neutrally-expected levels of diversity were calculated assuming no variance in female mating (e.g., $C^2_{f}=0$) and we considered a range of values for $\sigma_{m}$: 0 (no variance), 1/2 (Poisson/2) 1 (Poisson distribution of offspring), 2 (Poisson*2), and 3 (Poisson*3). Assuming the level of neutral polymorphism is given by the product of the mutation rate and the effective population size, $\theta=4N_{e}\mu$ \citep{watterson1975,kimura1984}, Figure 1 shows the predicted effective population size of autosomes and sex chromosomes as a function of the sex ratio and variance in male reproductive success. From these neutral predictions we calculated the normalized X/A and Y/A ratios of diversity as the basis for comparison with our empirical data.   Note that we have only considered variance in male reproductive success because this is most likely to occur in wind pollinated plant like \textit{R. hastatulus}, where males produce copious amounts of pollen and compete to fertilize uniovulate females.

@@

\subsection*{Simulations of positive and purifying selection}
To investigate whether our observed level of Y-chromosome diversity could be explained by selection, we conducted two sets of forward-time simulations using the software SFSCODE \citep{hernandez2008flexible}: (i) Purifying selection acting to remove deleterious mutations and linked neutral variation (i.e., background selection), and (ii) Positive selection on beneficial mutations reducing variation by selective sweeps (refs \X). For background selection simulations, we sampled selection coefficients from a gamma distributions, and investigated a range of values for the mean selection coefficient ($s$) by varying the scale parameter, $k$ (range \X). mutations rate $\mu=0.01$ as estimated in Hough et al. 2014,

%The effect of background selection is a function of the distribution of selection coefficients for new, deleterious mutations, and can be modeled by varying the mutation rate, the number of sites affected by selection (L), and the selection coefficient acting on new mutations (s) [21]. When evaluating models with different strengths of purifying selection, we assumed that selection coefficients for the nonsynonymous sites were drawn from a gamma distribution. Previous studies found this distribution to fit the observed autosomal frequency spectrum well [37,38,68,69], and there is little reason to believe that the shape of the gamma distribution varies across chromosomes.

To make the simulation output comparable to our data, we initialized simulations with our empirically estimated value of $\theta$ (from autosomes), and sampled 8 Y chromosomes per simulation from 24 populations. To consider the effects of selection against deleterious alleles (purifying selection), we sampled selective coefficients from a gamma distribution with shape parameter $k=0.1$, mean selection coefficient $s=-0.01$, neutral mutation rate $\mu = 0.005$, deleterious mutation rate $U=0.01$ (diploid). For positive selection simulations, we used \X [parameters here]. We ran 1000 replicates for each model and we calculated the average $\theta_{sim}$ over all simulation output to obtain the simulated Y/A ratio of neutral diversity under the both models (purifying and positive selection), taking into account unequal male and female population sizes. We compared simulated diversity output to our empirical data (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{sim}$ was not significantly different from $\theta_{observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supplementary Material.
